R graphics : ggplot2
--------------------------------------------------------

http://ggplot2.org

```{r}
library(ggplot2)
```

Improves the base and lattice systems by using a more formal underlying model.

Based on the concept of grammar of graphics, which describes the components that underline statistics graphics.

Components:
- **Data** we want to visualize, and **aestetic mappings** between this data and attribute we can perceive
- **Geoms** (geometric objects) what we see on the plot (point line...)
- **Stats** (statistical transformation) summarize, count observations, transformation
- **Scales**: legends, axes, etc. which make it possible to read the data from the graph
- **Coord**: a coordinate system
- **Facet**ing: how to break up data into subsets

### Functions


#### qplot

The equivalent of the general purpose plot() base function.

```{r, message=FALSE}
meatdata <- read.table(file="./meatdata.txt")
summary(meatdata)

#Default
qplot(year, amount, data=meatdata)

#One graph with points, colors function of meat origin
qplot(year, amount, data=meatdata, geom="point", colour=meat)
#One graph with lines, colors function of meat origin
qplot(year, amount, data=meatdata, geom="line", colour=meat)
#Multiple line graphs splitted by meat origin
qplot(year, amount, data=meatdata, geom="line", colour=meat, facets=meat~.)

#Other types of graphs
qplot(meat, amount, data=meatdata, geom="jitter")
qplot(meat, amount, data=meatdata, geom="boxplot")
qplot(meat, amount, data=meatdata, geom=c("boxplot", "jitter"), colour=meat)
qplot(amount, data=meatdata, geom="histogram", colour=meat, facets=meat~.)

```

When graphs are splitted by "facets", the group order depends on the corresponding factor's order (alphabetical, by default).


Other tests on the **Iris** data set.

```{r, message=FALSE}
data(iris)
qplot(x=Sepal.Length, y=Petal.Length, data=iris, geom="point")
qplot(x=Sepal.Length, y=Petal.Length, data=iris, geom="point", colour=Species)
qplot(x=Sepal.Length, y=Petal.Length, data=iris, geom="point", facets=Species~., colour=Species)
qplot(x=Sepal.Length, y=Petal.Length, data=iris, geom=c("smooth","point"), facets=Species~., colour=Species)

qplot(x=Species, y=Petal.Length, data=iris, geom=c("boxplot", "jitter"), colour=Species)
```
