R data structures
=============================================================

Any R object has 2 attributes: mode (type of data: logical, numeric, function, complex, character, raw, list,...) and class (label)


R storage mode
--------------

Obtained by the typeof() function
```{r}
f <- function() {}
f$a
## Error mentionning type 'closure'
```
logical, numeric (integer or double), function (closure or builtin or special)


Logical
----------

Easy to convert (and back) with numeric. But this easy conversion can cause problems.
```{r}
vector <- 1:10
vector[c(0,1)]
vector[c(F,T)]
```
Never use F and T shortcuts for False and TRUE.


Vector
-------------

Contain values of the **same mode**.
```{r}
x <- c(1,2,3,4); x
mode(x)
x <- c(1,2,TRUE,3);x
mode(x)
x <- c(1,2,"true", 4);x
mode(x)
```


Matrix
------------------

Can also access an element with one index
```{r}
a <- 1:30
attr(a, "dim") <- c(5,6)
class(a) <- "matrix"
a
a[11]
a[1,3]
```
As for vectors, all elements must be of the same mode. Insert a text, and all numeric will turn into characters.


List
----------------

Is a collection of objects with different modes. Access either by rank or by name.
```{r}
mylist <- list(ages=c(12,35,34,62), height=c(135,128,164,165), sex=c("M", "F", "M", "M"))
mylist[[1]]
class(mylist[[1]])
mylist[1]
class(mylist[1])
mylist$height
```



Data frames
------------------

List with each element has same number of elements.
```{r}
data <- as.data.frame(mylist)
data
class(data); mode(data)
data[1]
data[[1]]
data$h
```
Always use labels for accessing columns, **full labels** as partial labels imply R checking for all column names.

