Data manipulation/aggregation
================================================================================

Contingency table
--------------------------------------------------------------------------------

table()


"Apply" family
--------------------------------------------------------------------------------

### apply()

MAR=1 to loop on rows
MAR=2 to loop on columns

Other specialized function are often more efficient, like rowSums(matrix).

### lapply() 
to apply on the elements of a list

### sapply() 
to apply on the elements of a list and simplify the result to a vector if possible.

### tapply(data, factorForGroups, FUN) 
apply a function FUN on a matrix after aggregating data using a factor. Or one can use split(data[, "height"], data[ ,"sex"])


Aggregate
--------------------------------------------------------------------------------

aggregate(x, by, FUN) data.frame/**list**/function, works in a similar way to tapply() but:
- works on whole data.frame (multiple columns
- **only produces scalar summaries**

```{r}
data(iris)
head(iris, 3)
aggregate(iris[, 1:4], iris[5], FUN=mean)
## Note that iris[5] is a list (as iris["Species"]), 
## instead of iris[,5] which is a vector (as iris$Species)
```

Merge
--------------------------------------------------------------------------------

### merge()
Merge two data frames, by column or by row. By default the function finds all common column names and use them all.
Interesting parameters include :
'by' (by.x, by.y) to specify common columns to use for the merge
'all' (all.x, all.x) to specify if all lines should be kept (even the ones without correspondance in the other data frame). 
```{r}
x <- data.frame(index=c(2,3,1), fruit=c("orange", "lemon", "grapefruit"))
y <- data.frame(index=c(3,1,2,1), quantity=c(13, 6, 3, 5))
merge(x, y)
```

A good practice is to specify each time by which column you want to merge for x and y.

### union/intersect/setdiff/setequal

### match(newlist, reflist) 
a general way for finding common values


Split data according to groups information
--------------------------------------------------------------------------------

split() splits a vector (or rows of a data frame) into separate elements of a list, ready for further processing.

stack() transforms a matrix with headers (wide format), into one 2-columns data-frame with measurements/labels (long format).
Correlation need wide format, modeling need long format.

reshape() has more possibilities, and is consequently more complex.
